ANTS: Advanced Open-Source Tools for Normalization And Neuroanatomy

### 2009-06-26 Documentation/Tex/antstheory.tex
Changed bibliography path.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@26 fd2da0a4-d968-48df-9cb5-d8c496287cbf
ok - updated to compile
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@31 fd2da0a4-d968-48df-9cb5-d8c496287cbf
Changed the format to IEEE and updated the citations to reflect IEEE usage.  Also, changed the graphics path.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@46 fd2da0a4-d968-48df-9cb5-d8c496287cbf

% Template article for preprint document class `elsart'
% with harvard style bibliographic references
% SP 2006/04/26

\documentclass[preprint,twocolumn,authoryear,5p]{elsarticle}

@@

% Use the option doublespacing or reviewcopy to obtain double line spacing
 %\documentclass[doublespacing]{elsart}

% the natbib package allows both number and author-year (Harvard)
% style referencing;
\usepackage{natbib}

% if you use PostScript figures in your article
% use the graphics package for simple commands
% \usepackage{graphics}
% or use the graphicx package for more complicated commands
 \usepackage{graphicx}
% or use the epsfig package if you prefer to use the old commands
% \usepackage{epsfig}

% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb,amsmath}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{rotating}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{listing}						
\usepackage{listings}

\graphicspath{{/Users/nick/Documents/Academic/Gee/SubmittedPapers/JournalPapers/MedIA2009_BrianANTS/Figures/}
%{/Users/nick/Documents/Academic/Gee/SubmittedPapers/Conferences/ISBI2007/Images/}
%                          {/Users/nick/Documents/Academic/Gee/Projects/Helium-3/GrantSubmissions/R21/resplan/Images/}
%                          {/Users/nick/Documents/Academic/Gee/SubmittedPapers/JournalPapers/MRM2008/Figures/}
%                          {/Users/nick/Documents/Academic/Gee/SubmittedPapers/Conferences/ISBI2008/Figures/}
                          }

@@

\begin{frontmatter}

% Title, authors and addresses

% use the thanksref command within \title, \author or \address for footnotes;
% use the corauthref command within \author for corresponding author footnotes;
% use the ead command for the email address,
% and the form \ead[url] for the home page:
% \title{}
% \thanks[label1]{}
% \author{Name\corauthref{cor1}\thanksref{label2}}
% \ead{tustison@grasp.cis.upenn.edu}
% \ead[url]{home page}
% \thanks[label2]{}
% \corauth[cor1]{}
% \address{Address\thanksref{label3}}
% \thanks[label3]{}

\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
%\author[label1]{Brian B. Avants}
%\author[label1]{Nicholas J. Tustison}
%\author[label1]{Gang Song}
%\author[label1]{James C. Gee}
\author{Brian B. Avants}
\author{Nicholas J. Tustison}
\author{Gang Song}
\author{James C. Gee}% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
%\address[label1]{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
\address{Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, PA, USA}
%\address[label2]{Department of Mathematics, University of Pennsylvania, Philadelphia, PA, USA}

@@

\begin{keyword}
ANTS \sep image registration \sep nueroanatomy \sep neuroimaging \sep normalization \sep  open-source \sep point-set registration
% keywords here, in the form: keyword \sep keyword
% PACS codes here, in the form: \PACS code \sep code
\end{keyword}

\end{frontmatter}

@@

Despite the number of proposed algorithms, our limited assessment of published research mirrors the experience of many others who prefer a working paradigm of what has been referred to as {\em reproducible research}.  As described by Dr. Kovacevic ``[reproducible research] refers to the idea that, in `computational' sciences, the ultimate product is not a published paper but, rather, the entire environment used to produce the results in the paper (data, software,etc.).''  After an informal survey of 15 published papers, she found ``none had code available'' and ``in only about half the cases were the parameters [of the algorithm] specified'' (\cite{Kovacevic2006}).  Recent discussions within the computational sciences research community, particularly among advocates of ``open science,'' have also voiced similar concerns (\cite{Yoo2005,Ibanez2006}).   In this paper, we discuss our contribution to the open-source medical image analysis research community which we call ANTS (\underline{A}dvanced \underline{N}ormalization \underline{T}ool\underline{s}).  Built on an ITK framework, this software package comprises a suite of tools for image normalization and template building based on previously published research.

Perhaps the most persuasive evidence motivating the use of our contributions discussed in this paper is the recent outcome of a large-scale comparative image registration algorithm assessment (\cite{Klein2009}). In the largest evaluation study to date involving 14 popular non-linear registration algorithms, our Symmetric Normalization (SyN) transformation model (\cite{Avants2008}) discussed below, was consistently one of the top two performers across all tests.  Overlap and distance measures used for assessment employed three completely independent analysis methods (permutation tests, one-way ANOVA tests, and indifference zone ranking).  Unlike some of the other algorithms used in this brain registration evaluation study, all of our methods (not just SyN) are available as open-source.

@@

\section{Theoretical Overview of ANTS}
A useful classification schema of normalization techniques is based upon the following three principal components (\cite{Brown1992,Ibanez2002}):
\begin{itemize}
  \item the {\em transformation model},
  \item the {\em similarity (or correspondence) measures},  and
  \item the {\em optimization strategy}.
\end{itemize}
In general, image normalization is the process of finding the optimal transformation, $\phi$, within a specified transformation space which maps each $\mathbf{x}$ of image $\mathcal{I}(\mathbf{x})$ to a location in image $\mathcal{J}(\mathbf{z})$ such that a specified cost function, $\mathcal{C}$, describing the similarity between $\mathcal{I}$ and $\mathcal{J}$, is minimized.   A summary of available transformation models and similarity measures are provided in Table \ref{table:chart}.  Details are given in subsequent sections.

@@

A variety of transformation models have been proposed in the literature with varying degrees of freedom (illustrated in Figure \ref{fig:hierarchy}).  For deformable transformations, one approach is to optimize within the space of non-topology preserving, yet physically motivated transformations---an approach pioneered by Bajcsy (\cite{Bajcsy1989}).  Elastic and related models, such as HAMMER (\cite{Shen2002}), statistical parametric mapping (SPM) (\cite{Ashburner2000}), free-form deformations (FFD) (\cite{Rueckert1999}), and
Thirion's demons (\cite{Thirion1998}) operate in the space of vector fields, which does not preserve topology.  In other words, barring ad hoc constraints to prevent otherwise, these algorithms allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.

These topology concerns have led to current advocation within the computational anatomy research community of diffeomorphic mappings for transformation modeling.   Optimizing directly within this space has shown remarkable success in various computational anatomy studies involving longitudinal (\cite{Avants2007b,Fox2001}), functional (\cite{Miller2005}), and population data (\cite{Avants2007b}).  We include three such diffeomorphic algorithms in the ANTS toolbox based on previously reported research.

Regardless of current research trends, however, we recognize that selection of the transformation model is ultimately application-specific, that no single choice is optimal for all scenarios (\cite{Wolpert1997}), and, therefore, the transformation model must be chosen in a principled fashion.  In fact, several non-diffeomorphic algorithms performed quite well in Klein's comparative study of nonrigid registration algorithms (\cite{Klein2009}).  For this reason, we also include elastic-related methods for a panoply of transformation model options.

\begin{figure}
  \centering
    \begin{tabular}{c}
      \includegraphics[width=80mm]{hierarchy}
    \end{tabular}
  \caption{Diagrammatic illustration of the transformation model hierarchy where the encompassing transformation spaces are characterized by increasing degrees of freedom.  }
  \label{fig:hierarchy}
\end{figure}

@@

In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse (\cite{Ebin1970,Mumford1998}).  Modeling transformations with diffeomorphisms ensures certain desirable topological properties that cannot be guaranteed with other methods.

@@

Dupuis et al. (\cite{Dupuis1998}) motivated the usage of diffeomorphisms for CA  by showing that the variational form
\begin{align}\label{eq:D}
  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
\end{align}
represents a true mathematical metric between anatomical instances $\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the velocity field, $\boldsymbol{v}$.  An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.   Dupuis (\cite{Dupuis1998}) also showed that such a solution is guaranteed to be well-posed.  Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.  It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.%
\footnote{
It is important to note the similarity between the definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).  In this sense, the solution for Equation (\ref{eq:D}) is the geodesic diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the diffeomorphism, such that the shape distance, $D$, between $\mathcal{I}$ and $\mathcal{J}$ is minimized.
}

In most real-world applications, however, a diffeomorphic path connecting the anatomical instance $\mathcal{J}$ with $\mathcal{I}$ is non-existent (due, for example, to the photometric variation or the presence/absence of a tumor in neuroanatomical images).  Therefore, the following minimizing variational form is used for optimization in diffeomorphic normalization to accommodate inexact matching  (\cite{Dupuis1998,Miller2002})
\begin{align} \label{eq:lddmm}
  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2dt  + \lambda\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J}  || d\Omega \right\}.
\end{align}
The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) (\cite{Beg2005}) with the similarity metric, or data term, for LDDMM being the squared intensity difference with weight $\lambda$.

@@

Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
$\phi_1$ and $\phi_2$, \cite{Avants2007b} construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
\begin{align}\label{eq:diffs}
  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  = &
     \underset{\boldsymbol{v_1},\boldsymbol{v_2}}{\operatorname{argmin}}  
      \Bigg\{ \nonumber \\
      & \int_0^{0.5} \left( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   \right)dt  \nonumber \\
     &+ \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ \phi_1(\mathbf{x},0.5), \mathcal{J}\circ \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
\end{align}
The corresponding symmetric Euler-Lagrange equations are similar to (\cite{Miller2002}).  Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.  Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the solution strategy as Symmetric Normalization (SyN).

>>>>

%\documentclass[11pt,draftcls,onecolumn]{IEEEtran}
\documentclass[10pt,twocolumn]{IEEEtran}

@@

%% bare_jrnl.tex
%% V1.2
%% 2002/11/18
%% by Michael Shell
%% mshell@ece.gatech.edu
%%
%% NOTE: This text file uses MS Windows line feed conventions. When (human)
%% reading this file on other platforms, you may have to use a text
%% editor that can handle lines terminated by the MS Windows line feed
%% characters (0x0D 0x0A).
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.6b or later) with an IEEE journal paper.
%%
%% Support sites:
%% http://www.ieee.org
%% and/or
%% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/IEEEtran/
%%
%% This code is offered as-is - no warranty - user assumes all risk.
%% Free to use, distribute and modify.

% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
% *** with production work. IEEE's font choices can trigger bugs that do  ***
% *** not appear when using other class files.                            ***
% Testflow can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/testflow

% Note that the a4paper option is mainly intended so that authors in
% countries using A4 can easily print to A4 and see how their papers will
% look in print. Authors are encouraged to use U.S. letter paper when 
% submitting to IEEE. Use the testflow package mentioned above to verify
% correct handling of both paper sizes by the author's LaTeX system.
%
% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
% should be used if it is desired that the figures are to be displayed in
% draft mode.
%
% This example can be formatted using the peerreview
% (instead of journal) mode.
%\documentclass[11pt,draftcls,onecolumn]{IEEEtran}
\usepackage{amssymb,amsmath}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{rotating}
\usepackage{multirow,booktabs,ctable,array}
\usepackage{listing}						
\usepackage{listings}					
\graphicspath{./Figures/}

% If the IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it:
% \documentclass[journal]{../sty/IEEEtran}

% some very useful LaTeX packages include:

%\usepackage{cite}      % Written by Donald Arseneau
                        % V1.6 and later of IEEEtran pre-defines the format
                        % of the cite.sty package \cite{} output to follow
                        % that of IEEE. Loading the cite package will
                        % result in citation numbers being automatically
                        % sorted and properly "ranged". i.e.,
                        % [1], [9], [2], [7], [5], [6]
                        % (without using cite.sty)
                        % will become:
                        % [1], [2], [5]--[7], [9] (using cite.sty)
                        % cite.sty's \cite will automatically add leading
                        % space, if needed. Use cite.sty's noadjust option
                        % (cite.sty V3.8 and later) if you want to turn this
                        % off. cite.sty is already installed on most LaTeX
                        % systems. The latest version can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/cite/

%\usepackage{graphicx}  % Written by David Carlisle and Sebastian Rahtz
                        % Required if you want graphics, photos, etc.
                        % graphicx.sty is already installed on most LaTeX
                        % systems. The latest version and documentation can
                        % be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/required/graphics/
                        % Another good source of documentation is "Using
                        % Imported Graphics in LaTeX2e" by Keith Reckdahl
                        % which can be found as esplatex.ps and epslatex.pdf
                        % at: http://www.ctan.org/tex-archive/info/
% NOTE: for dual use with latex and pdflatex, instead load graphicx like:
%\ifx\pdfoutput\undefined
%\usepackage{graphicx}
%\else
%\usepackage[pdftex]{graphicx}
%\fi

% However, be warned that pdflatex will require graphics to be in PDF
% (not EPS) format and will preclude the use of PostScript based LaTeX
% packages such as psfrag.sty and pstricks.sty. IEEE conferences typically
% allow PDF graphics (and hence pdfLaTeX). However, IEEE journals do not
% (yet) allow image formats other than EPS or TIFF. Therefore, authors of
% journal papers should use traditional LaTeX with EPS graphics.
%
% The path(s) to the graphics files can also be declared: e.g.,
% \graphicspath{{../eps/}{../ps/}}
% if the graphics files are not located in the same directory as the
% .tex file. This can be done in each branch of the conditional above
% (after graphicx is loaded) to handle the EPS and PDF cases separately.
% In this way, full path information will not have to be specified in
% each \includegraphics command.
%
% Note that, when switching from latex to pdflatex and vice-versa, the new
% compiler will have to be run twice to clear some warnings.

%\usepackage{psfrag}    % Written by Craig Barratt, Michael C. Grant,
                        % and David Carlisle
                        % This package allows you to substitute LaTeX
                        % commands for text in imported EPS graphic files.
                        % In this way, LaTeX symbols can be placed into
                        % graphics that have been generated by other
                        % applications. You must use latex->dvips->ps2pdf
                        % workflow (not direct pdf output from pdflatex) if
                        % you wish to use this capability because it works
                        % via some PostScript tricks. Alternatively, the
                        % graphics could be processed as separate files via
                        % psfrag and dvips, then converted to PDF for
                        % inclusion in the main file which uses pdflatex.
                        % Docs are in "The PSfrag System" by Michael C. Grant
                        % and David Carlisle. There is also some information 
                        % about using psfrag in "Using Imported Graphics in
                        % LaTeX2e" by Keith Reckdahl which documents the
                        % graphicx package (see above). The psfrag package
                        % and documentation can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/psfrag/

%\usepackage{subfigure} % Written by Steven Douglas Cochran
                        % This package makes it easy to put subfigures
                        % in your figures. i.e., "figure 1a and 1b"
                        % Docs are in "Using Imported Graphics in LaTeX2e"
                        % by Keith Reckdahl which also documents the graphicx
                        % package (see above). subfigure.sty is already
                        % installed on most LaTeX systems. The latest version
                        % and documentation can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/subfigure/

%\usepackage{url}       % Written by Donald Arseneau
                        % Provides better support for handling and breaking
                        % URLs. url.sty is already installed on most LaTeX
                        % systems. The latest version can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/other/misc/
                        % Read the url.sty source comments for usage information.

%\usepackage{stfloats}  % Written by Sigitas Tolusis
                        % Gives LaTeX2e the ability to do double column
                        % floats at the bottom of the page as well as the top.
                        % (e.g., "\begin{figure*}[!b]" is not normally
                        % possible in LaTeX2e). This is an invasive package
                        % which rewrites many portions of the LaTeX2e output
                        % routines. It may not work with other packages that
                        % modify the LaTeX2e output routine and/or with other
                        % versions of LaTeX. The latest version and
                        % documentation can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/contrib/supported/sttools/
                        % Documentation is contained in the stfloats.sty
                        % comments as well as in the presfull.pdf file.
                        % Do not use the stfloats baselinefloat ability as
                        % IEEE does not allow \baselineskip to stretch.
                        % Authors submitting work to the IEEE should note
                        % that IEEE rarely uses double column equations and
                        % that authors should try to avoid such use.
                        % Do not be tempted to use the cuted.sty or
                        % midfloat.sty package (by the same author) as IEEE
                        % does not format its papers in such ways.

%\usepackage{amsmath}   % From the American Mathematical Society
                        % A popular package that provides many helpful commands
                        % for dealing with mathematics. Note that the AMSmath
                        % package sets \interdisplaylinepenalty to 10000 thus
                        % preventing page breaks from occurring within multiline
                        % equations. Use:
%\interdisplaylinepenalty=2500
                        % after loading amsmath to restore such page breaks
                        % as IEEEtran.cls normally does. amsmath.sty is already
                        % installed on most LaTeX systems. The latest version
                        % and documentation can be obtained at:
                        % http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/

% Other popular packages for formatting tables and equations include:

%\usepackage{array}
% Frank Mittelbach's and David Carlisle's array.sty which improves the
% LaTeX2e array and tabular environments to provide better appearances and
% additional user controls. array.sty is already installed on most systems.
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/required/tools/

% Mark Wooding's extremely powerful MDW tools, especially mdwmath.sty and
% mdwtab.sty which are used to format equations and tables, respectively.
% The MDWtools set is already installed on most LaTeX systems. The lastest
% version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/mdwtools/

% V1.6 of IEEEtran contains the IEEEeqnarray family of commands that can
% be used to generate multiline equations as well as matrices, tables, etc.

% Also of notable interest:

% Scott Pakin's eqparbox package for creating (automatically sized) equal
% width boxes. Available:
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/eqparbox/

% Notes on hyperref:
% IEEEtran.cls attempts to be compliant with the hyperref package, written
% by Heiko Oberdiek and Sebastian Rahtz, which provides hyperlinks within
% a document as well as an index for PDF files (produced via pdflatex).
% However, it is a tad difficult to properly interface LaTeX classes and
% packages with this (necessarily) complex and invasive package. It is
% recommended that hyperref not be used for work that is to be submitted
% to the IEEE. Users who wish to use hyperref *must* ensure that their
% hyperref version is 6.72u or later *and* IEEEtran.cls is version 1.6b
% or later. The latest version of hyperref can be obtained at:
%
% http://www.ctan.org/tex-archive/macros/latex/contrib/supported/hyperref/
%
% Also, be aware that cite.sty (as of version 3.9, 11/2001) and hyperref.sty
% (as of version 6.72t, 2002/07/25) do not work optimally together.
% To mediate the differences between these two packages, IEEEtran.cls, as
% of v1.6b, predefines a command that fools hyperref into thinking that
% the natbib package is being used - causing it not to modify the existing
% citation commands, and allowing cite.sty to operate as normal. However,
% as a result, citation numbers will not be hyperlinked. Another side effect
% of this approach is that the natbib.sty package will not properly load
% under IEEEtran.cls. However, current versions of natbib are not capable
% of compressing and sorting citation numbers in IEEE's style - so this
% should not be an issue. If, for some strange reason, the user wants to
% load natbib.sty under IEEEtran.cls, the following code must be placed
% before natbib.sty can be loaded:
%
% \makeatletter
% \let\NAT@parse\undefined
% \makeatother
%
% Hyperref should be loaded differently depending on whether pdflatex
% or traditional latex is being used:
%
%\ifx\pdfoutput\undefined
%\usepackage[hypertex]{hyperref}
%\else
%\usepackage[pdftex,hypertexnames=false]{hyperref}
%\fi
%
% Pdflatex produces superior hyperref results and is the recommended
% compiler for such use.

% *** Do not adjust lengths that control margins, column widths, etc. ***
% *** Do not use packages that alter fonts (such as pslatex).         ***
% There should be no need to do such things with IEEEtran.cls V1.6 and later.

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

@@

%
% paper title
\title{ANTS:  Advanced Open-Source Tools for Normalization And Neuroanatomy}
%
%
% author names and IEEE memberships
% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
\author{Brian B. Avants, Nicholas J. Tustison, Gang Song, and James C. Gee
%               Penn Image Computing and Science Laboratory \\
%               University of Pennsylvania \\
%               3600 Market Street, Suite 370 \\
%               Philadelphia,  PA  19104 \\
%               contact email:  tustison@grasp.cis.upenn.edu
%\thanks{Manuscript received XXX; revised XXX.
 %       }% <-this % stops a space
}

% note the % following the last \IEEEmembership and also the first \thanks - 
% these prevent an unwanted space from occurring between the last author name
% and the end of the author line. i.e., if you had this:
% 
% \author{....lastname \thanks{...} \thanks{...} }
%                     ^------------^------------^----Do not want these spaces!
%
% a space would be appended to the last name and could cause every name on that
% line to be shifted left slightly. This is one of those "LaTeX things". For
% instance, "A\textbf{} \textbf{}B" will typeset as "A B" not "AB". If you want
% "AB" then you have to do: "A\textbf{}\textbf{}B"
% \thanks is no different in this regard, so shield the last } of each \thanks
% that ends a line with a % and do not let a space in before the next \thanks.
% Spaces after \IEEEmembership other than the last one are OK (and needed) as
% you are supposed to have spaces between the names. For what it is worth,
% this is a minor point as most people would not even notice if the said evil
% space somehow managed to creep in.
%
% The paper headers
\markboth{IEEE Transactions on Medical Imaging,~Vol.~X, No.~XX,~November~200X}{Tustison\MakeLowercase{\textit{et al.}}: Advanced Normalization Tools}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's name in ***
% *** the headers of peer review papers.                                   ***

% If you want to put a publisher's ID mark on the page
% (can leave text blank if you just want to see how the
% text height on the first page will be reduced by IEEE)
%\pubid{0000--0000/00\$00.00~\copyright~2002 IEEE}

% use only for invited papers
%\specialpapernotice{(Invited Paper)}

% make the title area

\maketitle

@@

\IEEEkeywords{
ANTS, image registration, nueroanatomy, neuroimaging, normalization, open-source, point-set registration
}

\IEEEpeerreviewmaketitle

@@

Despite the number of proposed algorithms, our limited assessment of published research mirrors the experience of many others who prefer a working paradigm of what has been referred to as {\em reproducible research}.  As described by Dr. Kovacevic ``[reproducible research] refers to the idea that, in `computational' sciences, the ultimate product is not a published paper but, rather, the entire environment used to produce the results in the paper (data, software,etc.).''  After an informal survey of 15 published papers, she found ``none had code available'' and ``in only about half the cases were the parameters [of the algorithm] specified'' \cite{Kovacevic2006}.  Recent discussions within the computational sciences research community, particularly among advocates of ``open science,'' have also voiced similar concerns \cite{Yoo2005,Ibanez2006}.   In this paper, we discuss our contribution to the open-source medical image analysis research community which we call ANTS (\underline{A}dvanced \underline{N}ormalization \underline{T}ool\underline{s}).  Built on an ITK framework, this software package comprises a suite of tools for image normalization and template building based on previously published research.

Perhaps the most persuasive evidence motivating the use of our contributions discussed in this paper is the recent outcome of a large-scale comparative image registration algorithm assessment \cite{Klein2009}. In the largest evaluation study to date involving 14 popular non-linear registration algorithms, our Symmetric Normalization (SyN) transformation model \cite{Avants2008} discussed below, was consistently one of the top two performers across all tests.  Overlap and distance measures used for assessment employed three completely independent analysis methods (permutation tests, one-way ANOVA tests, and indifference zone ranking).  Unlike some of the other algorithms used in this brain registration evaluation study, all of our methods (not just SyN) are available as open-source.

@@

\section{Theoretical Overview of ANTS}
A useful classification schema of normalization techniques is based upon the following three principal components \cite{Brown1992,Ibanez2002}:
\begin{itemize}
  \item the {\em transformation model},
  \item the {\em similarity (or correspondence) measures},  and
  \item the {\em optimization strategy}.
\end{itemize}
In general, image normalization is the process of finding the optimal transformation, $\phi$, within a specified transformation space which maps each $\mathbf{x}$ of image $\mathcal{I}(\mathbf{x})$ to a location in image $\mathcal{J}(\mathbf{z})$ such that a specified cost function, $\mathcal{C}$, describing the similarity between $\mathcal{I}$ and $\mathcal{J}$, is minimized.   A summary of available transformation models and similarity measures are provided in Table \ref{table:chart}.  Details are given in subsequent sections.

@@

A variety of transformation models have been proposed in the literature with varying degrees of freedom (illustrated in Figure \ref{fig:hierarchy}).  For deformable transformations, one approach is to optimize within the space of non-topology preserving, yet physically motivated transformations---an approach pioneered by Bajcsy \cite{Bajcsy1989}.  Elastic and related models, such as HAMMER \cite{Shen2002}, statistical parametric mapping (SPM) \cite{Ashburner2000}, free-form deformations (FFD) \cite{Rueckert1999}, and
Thirion's demons \cite{Thirion1998} operate in the space of vector fields, which does not preserve topology.  In other words, barring ad hoc constraints to prevent otherwise, these algorithms allow the topology to change in an uncontrolled way which makes the deformable mappings difficult to interpret in functional or anatomical studies.

These topology concerns have led to current advocation within the computational anatomy research community of diffeomorphic mappings for transformation modeling.   Optimizing directly within this space has shown remarkable success in various computational anatomy studies involving longitudinal \cite{Avants2007b,Fox2001}, functional \cite{Miller2005}, and population data \cite{Avants2007b}.  We include three such diffeomorphic algorithms in the ANTS toolbox based on previously reported research.

Regardless of current research trends, however, we recognize that selection of the transformation model is ultimately application-specific, that no single choice is optimal for all scenarios \cite{Wolpert1997}, and, therefore, the transformation model must be chosen in a principled fashion.  In fact, several non-diffeomorphic algorithms performed quite well in Klein's comparative study of nonrigid registration algorithms \cite{Klein2009}.  For this reason, we also include elastic-related methods for a panoply of transformation model options.

\begin{figure}
  \centering
    \begin{tabular}{c}
      \includegraphics[width=80mm]{Figures/hierarchy}
    \end{tabular}
  \caption{Diagrammatic illustration of the transformation model hierarchy where the encompassing transformation spaces are characterized by increasing degrees of freedom.  }
  \label{fig:hierarchy}
\end{figure}

@@

In contrast to many transformation models which reside in the domain of vector spaces, a diffeomorphism is a differentiable mapping with a differentiable inverse \cite{Ebin1970,Mumford1998}.  Modeling transformations with diffeomorphisms ensures certain desirable topological properties that cannot be guaranteed with other methods.

@@

Dupuis et al. \cite{Dupuis1998} motivated the usage of diffeomorphisms for CA  by showing that the variational form
\begin{align}\label{eq:D}
  D(\mathcal{I}, \mathcal{J}) =  \int_0^1 ||L\boldsymbol{v}||^2 dt , \,\, \mathcal{I}(\phi(\mathbf{x}, 1)) = \mathcal{J}(\mathbf{z})
\end{align}
represents a true mathematical metric between anatomical instances $\mathcal{I}$ and $\mathcal{J}$ given an appropriate norm, $L$, on the velocity field, $\boldsymbol{v}$.  An optimal solution, $\boldsymbol{v}^*$, minimizes the metric $D(\mathcal{I}, \mathcal{J})$ with respect to $L$.   Dupuis \cite{Dupuis1998} also showed that such a solution is guaranteed to be well-posed.  Intuitively, Equation (\ref{eq:D}) provides a sense of distance between two anatomical shapes.  It also illustrates that the optimal diffeomorphic solution is analogous to finding the geodesic curve between two points in a curved space.%
\footnote{
It is important to note the similarity between the definition of curve length, $\int ||\mathcal{C}'(t)||dt$, for the parametric curve $\mathcal{C}(t)$ and Equation (\ref{eq:D}).  In this sense, the solution for Equation (\ref{eq:D}) is the geodesic diffeomorphism, where $\boldsymbol{v}$ is the tangent vector of the diffeomorphism, such that the shape distance, $D$, between $\mathcal{I}$ and $\mathcal{J}$ is minimized.
}

In most real-world applications, however, a diffeomorphic path connecting the anatomical instance $\mathcal{J}$ with $\mathcal{I}$ is non-existent (due, for example, to the photometric variation or the presence/absence of a tumor in neuroanatomical images).  Therefore, the following minimizing variational form is used for optimization in diffeomorphic normalization to accommodate inexact matching  \cite{Dupuis1998,Miller2002}
\begin{align} \label{eq:lddmm}
  \boldsymbol{v}^* = \underset{\boldsymbol{v}}{\operatorname{argmin}}  \left\{ \int_0^1  ||L\boldsymbol{v}||^2dt  + \lambda\int_{\Omega} || \mathcal{I} \circ \phi(\mathbf{x},1) - \mathcal{J}  || d\Omega \right\}.
\end{align}
The Euler-Lagrange equations characterizing the optimizing time-varying velocity field, $\boldsymbol{v}^*$, were derived in \cite{Miller2002} and later used in formulating the gradient-descent optimization scheme known as {\em large deformation diffeomorphic metric-matching} (LDDMM) \cite{Beg2005} with the similarity metric, or data term, for LDDMM being the squared intensity difference with weight $\lambda$.

@@

Exploiting the fact that the diffeomorphism, $\phi$, can be decomposed into two components 
$\phi_1$ and $\phi_2$, \cite{Avants2007b} construct a {\em symmetric} alternative to Equation (\ref{eq:diff}).  This leads to the symmetric variant of Equation (\ref{eq:diff})
\begin{align}\label{eq:diffs}
  \{\boldsymbol{v}_1^*,\boldsymbol{v}_2^*\}  = &
     \underset{\boldsymbol{v_1},\boldsymbol{v_2}}{\operatorname{argmin}}  
      \Bigg\{ \nonumber \\
      & \int_0^{0.5} \left( ||L\boldsymbol{v}_1(x,t)||^2 + ||L\boldsymbol{v}_2(x,t)||^2   \right)dt  \nonumber \\
     &+ \lambda\int_{\Omega}\Pi_{\sim}\left( \mathcal{I} \circ \phi_1(\mathbf{x},0.5), \mathcal{J}\circ \phi_2(\mathbf{x},0.5)\right)d\Omega\Bigg\}.
\end{align}
The corresponding symmetric Euler-Lagrange equations are similar to \cite{Miller2002}.  Finding $\boldsymbol{v_1}^*$ minimizes the variational energy from $t = 0$ whereas $\boldsymbol{v_2}^*$ minimizes from $t = 1$.  Thus, gradient-based iterative convergence deforms $\mathcal{I}$ and $\mathcal{J}$ along the geodesic diffeomorphism, $\phi$, to a fixed  point midway (intuited by the notion of shape distance) between $\mathcal{I}$ and $\mathcal{J}$ thus motivating the denotation of the solution strategy as Symmetric Normalization (SyN).

Other diffeomorphic flavors have since been reported in the research literature (e.g. DARTEL \cite{Ashburner2007} and Diffeomorphic Demons \cite{Vercauteren2007,Vercauteren2009}.  We include three diffeomorphic transformation models for parameterizing $\phi(\cdot)$.  These include Geodesic SyN, Greedy SyN, and exponential mapping.  As summarized in Table \ref{table:chart}, each of these transformation models can utilize a host of similarity measures both individually and in mutual combination.
